---
title: "crayon colors, simplified"
author: Karl
date: '2017-05-02'
categories: ['R']
tags: ['R', 'data visualization']
---

```{r options, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width=7.2)
```

Back in 2014, I'd grabbed the [Crayola crayon colors from
Wikipedia](https://en.wikipedia.org/wiki/List_of_Crayola_crayon_colors)
and made them available as part of my [R/broman](https://github.com/kbroman/broman)
package.

(I [wrote about it on my wordpress blog](https://kbroman.wordpress.com/2014/05/07/crayon-colors-in-r/),
and then
[again](https://kbroman.wordpress.com/2014/05/08/two-more-points-about-crayon-colors/)
and
[again](https://kbroman.wordpress.com/2014/05/09/further-points-on-crayon-colors/).)

But using my package to get the crayon colors can be rather tedious.

It's easy to view the available colors.

```{r plot_crayons, fig.height=6, fig.width=10.5, out.width=700}
library(broman)
plot_crayons()
```

But grabbing a color, say "Purple Mountain's Majesty", requires rather
verbose code.

```{r purple_mountains_majesty}
brocolors("crayons")["Purple Mountain's Majesty"]
```

This morning, it was just too much for me. So I wrote a simpler
function `crayons()`, which accepts partial matches, using `grep`.
So you can write things like:

```{r purple_m}
crayons("purple m")
crayons("tickle")
crayons(c("purple m", "tickle"))
crayons("purple m", "tickle")
```

Why didn't I think of this before?

The new version of the package is not yet on
[CRAN](https://cran.r-project.org), but you can install it from
[GitHub](https://github.com/kbroman/broman) using
[devtools](https://github.com/hadley/devtools):

```{r install_broman, eval=FALSE}
devtools::install_github("kbroman/broman")
```

Or you can ignore the rest of the package and just [grab the
code](https://github.com/kbroman/broman/blob/master/R/brocolors.R).
