---
title: 'miniCRAN'
author: Karl
date: '2017-05-23T23:42:00-05:00'
categories: [R]
tags: ['R']
---

```{r knitr_options, include=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```

Next month, I'm teaching a workshop on [R/qtl](https://rqtl.org) and
[R/qtl2](https://kbroman.org/qtl2) at the
[joint Complex Trait Community and Rat Genomics Community meeting
(CTC-RG2017)](http://www.complextrait.org/ctc2017) meeting in Memphis.

I was chatting with [&#346;aunak Sen](http://www.senresearch.org/)
about preparations: the main problem is installation of R/qtl2, which
is split into multiple R packages, and I've not yet placed them on
[CRAN](https://cran.r-project.org). And they're a bit of a pain to
install from GitHub (people need the proper development tools, and
compilation of the C++ code is painfully slow).

Should we have participants work on a server?
Should we set up [RStudio](https://rstudio.com) in
[Docker](https://www.docker.com/) containers on [Digital
Ocean](https://www.digitalocean.com)?

&#346;aunak said, "What we need is a mini-CRAN." A google search
brought us to the package
[miniCRAN](https://cran.r-project.org/package=miniCRAN)
from [Revolution Analytics](https://blog.revolutionanalytics.com/),
which looks to be the perfect solution.

As it turns out, to make your own
[CRAN](https://cran.r-project.org) you "just" need to put all of the
requisite packages (source plus windows and mac binaries) into a
specific directory structure, with each directory having special
files (`PACKAGE`, `PACKAGE.gz`, and `PACKAGE.rds`) that describe the
contents.

The main trick is ensuring that you have all of the required
dependencies, and that's where
[miniCRAN](https://cran.r-project.org/package=miniCRAN) is especially
helpful.

[miniCRAN](https://cran.r-project.org/package=miniCRAN) has several
useful vignettes, plus a nice [blog
post](https://blog.revolutionanalytics.com/2014/10/introducing-minicran.html)
that explains the goal and basic use.

After installing miniCRAN, the first thing was to define the set of
packages that my [R/qtl2](https://kbroman.org/qtl2) packages depend on,
and then use `miniCRAN::pkgDep` to find all of the packages that those
depend on.

```{r pkg_dependencies}
library(miniCRAN)
repo <- "https://cran.rstudio.com"
pkgs <- c("Rcpp", "RcppEigen", "yaml", "jsonlite", "data.table",
          "knitr", "rmarkdown", "qtl", "qtlcharts")
pkgList <- pkgDep(pkgs, repos=repo, type="source", suggests=FALSE)
```

Here `pkgs` is a vector containing the main packages that we'll need
(other than the R/qtl2 packages), and `pkgList` is a longer vector
that includes all packages that _they_ depend on.

We then use `miniCRAN::makeRepo` to download all of those packages and
place them in the required directory structure, with the main
directory being `qtl2cran`:

```{r make_repo}
makeRepo(pkgList, path="qtl2cran", repos=repo,
         type=c("source", "win.binary", "mac.binary.el-capitan"))
```

The only issue here is that the version of miniCRAN on CRAN doesn't
handle `mac.binary.el-capitan` properly, nor does the [version on
GitHub](https://github.com/revolutionAnalytics/miniCRAN). But there's
a [pull
request](https://github.com/RevolutionAnalytics/miniCRAN/pull/84) from
[Vincent Nijs](https://github.com/vnijs) that fixes that, so I used
`devtools::install_github("vnijs/miniCRAN")` to use his version.

I then needed to add my [R/qtl2](https://kbroman.org/qtl2) packages:

- source `.tar.gz` files in `qtl2cran/src/contrib`
- Windows `.zip` files in `qtl2cran/bin/windows/contrib/3.4`
- Mac `.tgz` files in `qtl2cran/bin/macosx/el-capitan/contrib/3.4`

I'm working on a Mac (with OS X Sierra) so I could create the Mac
binaries by installing the packages, going to the package directory,
and using for example

```
tar czvf qtl2geno_0.5-21.tgz qtl2geno
```

For the windows binaries, I untarred the source files on my [windows
laptop](https://twitter.com/kwbroman/status/540250068360916994) and
used, for example:

```
R CMD INSTALL --build qtl2geno
```

This makes the `qtl2geno_0.5-21.zip` file.

The last step is to create those `PACKAGE*` files. A bit of poking
around in the [miniCRAN
source](https://github.com/RevolutionAnalytics/miniCRAN) pointed me to
`tools::write_PACKAGES()`, and so I used:

```{r write_PACKAGES}
library(tools)
write_PACKAGES("qtl2cran/src/contrib", type="source")
write_PACKAGES("qtl2cran/bin/windows/contrib/3.4", type="win.binary")
write_PACKAGES("qtl2cran/bin/macosx/el-capitan/contrib/3.4",
               type="mac.binary")
```

I then put them on my website at UW-Madison plus at
[rqtl.org](https://rqtl.org), so now people can install
[R/qtl2](https://kbroman.org) using either

```{r cran_rqtl}
install.packages(paste0("qtl2", c("geno", "scan", "plot", "convert")),
                 repos="https://rqtl.org/qtl2cran")
```

or

```{r cran_biostat}
install.packages(paste0("qtl2", c("geno", "scan", "plot", "convert")),
                 repos="https://www.biostat.wisc.edu/~kbroman/qtl2cran")
```

We just need to make sure that everyone has R version 3.4.0. And the
Mac users will need a relatively recent version of Mac OS X (El Capitan
or Sierra).

Long term, I need to get my packages on the proper
[CRAN](https://cran.r-project.org), but I really don't want to have to
deal with that in the next three weeks, and so I'm glad for
[miniCRAN](https://cran.r-project.org/package=miniCRAN), as it helped
me towards this short-term solution.
